Stochastic blockmodeling of relational event dynamics

### 2012-06-01 docs/intro.tex
Move some citations around.

In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.  While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.  Consider, e.g., Figure~\ref{fig:introexample}.  In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.  As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.  By contrast, consider the interaction patterns shown in the top right panel.  Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.

@@

Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
In this way one can compare theories about the underlying processes and make predictions about future data conditioned on the past.

>>>>

In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.  While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.  Consider, e.g., Figure~\ref{fig:introexample}.  In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.  As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.  By contrast, consider the interaction patterns shown in the top right panel.  Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.

@@

Borrowing from the intuition of stochastic blockmodels, we propose a novel continuous-time model of network-based event sequences where latent clusters of nodes share similar patterns of interaction.
Our approach employs a flexible framework for  specifying how the process depends on the previous history of events \cite{AalenOddO.2008, Butts2008}.
In this way one can compare theories about the underlying processes and make predictions about future data conditioned on the past.
{\color{red} ...and simultaneously adjust for heterogeneity in these
  processes?}

### 2012-06-01 docs/model.tex
Move some citations around.

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
\caption{Intensities for two dyads}
\end{subfigure}
\caption[]{Illustration of relational event data and the assumptions of the model.  (a) A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$. (b) The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{3,4}(t)$ (dotted).  We assume $\lambda_{3,4}$ can only change after events sent or received by node 3. }
\label{fig:example}
\end{figure}

@@

\subsection{Related work}
The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2010,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
We instead use latent variables to allow for heterogeneity in intensities across possible events.

>>>>

% labels
\node at (-.5,2) {$\lambda_{1,2}$};
\node at (-.5,1) {$\lambda_{3,4}$};
\node at (1,-.3) {$t_1$};
\node at (3,-.3) {$t_2$};
\end{tikzpicture}
\caption{Intensities for two dyads}
\end{subfigure}
\caption[]{Illustration of relational event data and the assumptions of the model.  (a) A sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$. (b) The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{3,4}(t)$ (dotted).  We assume $\lambda_{3,4}(t)$ can only change after events sent or received by node 3. }
\label{fig:example}
\end{figure}

@@

\subsection{Related work}
Stochastic blockmodels have been extended to longitudinal network data
involving a sequence of networks occurring at discrete times
\cite{Ishiguro2010, Rodriguez2011}, while we focus on methods for
analyzing a stream of relational events. 
The proposed model is an extension of recent work on modeling event-based social network data using an event history approach \cite{Butts2008,Brandes2009,Perry2011,Stadtfeld2011,Opsahl2011,Vu2011,Vu2011a}. % \cite{AalenOddO.2008} 
Relational event models such as these require a knot at each observed event, while other approaches such as \cite{Gunawardana2011} learn the regions where an intensity is constant.
By using decision trees,  \cite{Gunawardana2011} also allow for a nonlinear relationship between statistics and intensity functions.  
We instead use latent variables to allow for heterogeneity in intensities across the set of possible events.
